# Some basic examples {#VI}

This chapter presents some examples of text output generated by the `VI()` command of the `BrailleR` package. These examples generate output that is displayed in the R session just like any output from standard R commands. Please note however that not all `VI()` commands behave in this fashion; some more advanced uses of `VI()` are discussed in a later chapter. 

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
library(BrailleR)
```

You will need the `BrailleR` package to be ready for use to follow along with the examples in this chapter. Do this by issuing the command `library(BrailleR)` now.



## Histograms

The first and most commonly used example demonstrating the value of the BrailleR  package to a blind user is the creation of a histogram.

```{r hist, fig.cap="A histogram of 1000 random values from a normal distribution"}
x=rnorm(1000)
VI(hist(x))
```

The `VI()` command actually calls the `VI.histogram()` command as the `hist()` command creates an object of class "histogram". This means we can tailor the output to the information needed for any histogram created using the `hist()` command.

### Important features

The `VI()` command has added to the impact of issuing the `hist()` command as the actual graphic is generated for the sighted audience. The blind user can read from the text description so that they can interpret the information that the histogram offers the sighted world.

The above example showed the standard implementation of the `hist()` function. The `hist()` function of the `graphics` package does not store the additional arguments that improve the visual attractiveness. The solution (perhaps temporary) is to mask the original function with one included in the `BrailleR` package that calls the `graphics` package function, and then adds extra detail for any added plotting arguments.

This is best illustrated using the example included in the `BrailleR::hist()` function.

```{r BrailleRHistExample}
example(hist)
```


### Warning

The `VI()` function is partially reliant on the use of the `hist()` function that is included in the `BrailleR` package. If a histogram is created using a command that directly links to the original `hist()` command found in the `graphics` package, then the `VI()` command's output will not be as useful to the blind user. This mainly affects the presentation of the title and axis labels; it should not affect the details of the counts etc. within the histogram itself.

This behaviour could arise if the histogram is sought indirectly. If for example, a function offers (as a side effect) to create a histogram, the author of the function may have explicitly stated use of the `hist()` function from the `graphics` package using `graphics::hist()` instead of `hist()`. Use of `graphics::hist()` will bypass the `BrailleR::hist()` function that the `VI()` command needs. This should not create error messages, but may result in some strange and possibly undesirable output.



## Basic numerical summaries

The standard presentation of a summary of a data frame where each variable is given its own column is difficult for a screen reader user to read as the processing of information is done line by line. For example:

```{r UseSummary}
summary(airquality)
```

The `VI()` command actually calls the `VI.data.frame()` command. It then processes each variable one by one so that the results are printed  variable by variable instead of summary statistic by summary statistic. For example:
```{r UseVI}
VI(airquality)
```



### Important features

Note that in this case, the blind user could choose to present the summary of each variable as generated by the `VI()` command, or the output from the standard `summary()` command. There is no difference in the information that is ultimately presented in this case.


## BrailleR commands used in this chapter

The only explicit command from the `BrailleR` package used in this chapter was the `VI()` command.
