# The ggplot world and BrailleR {#GGPlot}


The use of the ggplot style of graph production has increased markedly since its inception. The grammar of graphics as seen in the R code used to create the extremely wide range of graphs is seldom human-interpretable with ease. Creation of suitable support functionality via the `VI()` command is very definitely required. An initial attempt to extract any useful infromation from these graphs was contributed to the `BrailleR` package by Tony Hirst. Significant improvement has been made by Debra Warren as part of her postgraduate work under the supervision of Paul Murrell at the University of Auckland. Much of what is displayed in this chapter is only worth offering because of Debra's work and the interactions had between her, Paul and I in the second half of 2017.


```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=7, fig.height=5, comment="")
library(BrailleR)
```



N.B.  the commands here are exact copies of the commands presented in @Wickham2009ggplot2 or some minor alterations to them; any changes will be explicitly noted.    All plots are created using the figure numbers from @Wickham2009ggplot2 or the page numbers if no figure number was given. They are then investigated using the `VI()` command from the `BrailleR` package.


You will need some additional packages to the `BrailleR` package to be ready for use to follow along with the examples in this chapter. Do this by issuing the commands:

```{r GetLibraries}     
library(BrailleR)   
library(ggplot2)   
library(magrittr)
```

Note that one data set used for these examples is created by @Wickham2009ggplot2 while the others are included in the `ggplot2` package.
```{r makeData}
set.seed(1410)
dsmall <- diamonds[sample(nrow(diamonds), 100),]   
```  


One important note for the graphs in this chapter is the difference in the way they appear here,  as compared to the original figures of @Wickham2009ggplot2 where differing height and width parameters have been set for each graph. For example,  in the following graph, the points are smaller than in the original figure, and the aspect ratio is slightly different. The consequence is that this graph looks less cluttered than does the original.

```{r p11a, fig.cap="First graph on page 11 of @Wickham2009ggplot2"}     
p11a = qplot(carat, price, data = diamonds)   
p11a
VI(p11a)  
```  

Note that unlike some other ways the `VI()` command has worked, the graph was not created by the nesting of the call to render the graph when nested inside the `VI()` command. In all the examples that follow, I use the pipe `%>%` operator from the `magrittr` package to push the graph into the `VI()` function.

```{r fig2-2a, fig.cap="Left pane of Figure 2.2 of @Wickham2009ggplot2"}
fig2.2a = qplot(carat, price, data = dsmall, colour = color)
fig2.2a %>% VI()   
fig2.2a
```

We haven't been able to tell what exact colour was used in the @Wickham2009ggplot2 rendering of this graph,  but there has obviously been some minor alteration of the colour palette being used by the `ggplot2` package.

```{r fig2-2b, fig.cap="Right pane of Figure 2.2"}
fig2.2b = qplot(carat, price, data = dsmall, shape = cut)    
fig2.2b %>% VI()   
fig2.2b
```  

To get semi-transparent points:   
```{r fig2-3b, fig.cap="Middle pane from Figure 2.3"}     
fig2.3b = qplot(carat, price, data = diamonds, alpha = I(1/100))    
fig2.3b
fig2.3b %>% VI()   
```  

To add a smoother (default is loess for n<1000):   
```{r fig2-4a, fig.cap="Left pane of Figure 2.4"}     
fig2.4a = qplot(carat, price, data = dsmall, geom = c("point", "smooth"))    
fig2.4a
fig2.4a %>% VI()   
```  



## Plotting a  continuous variable against a categorical variable  

```{r fig2-8a, fig.cap="Left pane of Figure 2.8"}     
fig2.8a  = qplot(color, price / carat, data = diamonds, geom = "jitter")
fig2.8a
fig2.8a %>% VI()   
```

```{r fig2-8b}
fig2.8b = qplot(color, price / carat, data = diamonds, geom = "boxplot")    
fig2.8b
fig2.8b %>% VI()   
```  

When seeking to use shading or opaqueness to describe the density of the points, the fact the size of the points has an impact on the opaqueness is not realised by `BrailleR`. 
```{r fig2-9b, fig.cap="Middle pane of Figure 2.9"}     
fig2.9b = qplot(color, price / carat, data = diamonds, geom = "jitter", alpha = I(1 / 50))    
fig2.9b
fig2.9b %>% VI()   
```


### univariate plots   


```{r fig2-10a, fig.cap="Left pane of Figure 2.10"}     
fig2.10a = qplot(carat, data = diamonds, geom = "histogram")    
fig2.10a
fig2.10a %>% VI()   
```
Warning: This figure does look different to the original in @Wickham2009ggplot2 ins spite of using the same code and same data.

```{r fig2-10b, fig.cap="Right pane of Figure 2.10"}     
fig2.10b = qplot(carat, data = diamonds, geom = "density")    
fig2.10b
fig2.10b %>% VI()   
```


```{r fig2-11c, fig.cap="Right pane of Figure 2.11"}     
fig2.11c = qplot(carat, data = diamonds, geom = "histogram", binwidth = 0.01, xlim = c(0,3))    
fig2.11c
fig2.11c %>% VI()   
```  

The data is separated by implication in the following graphs. The legend is automatically generated and has altered in appearance since the original was prodcued in @Wickham2009ggplot2.

```{r fig2-12a, fig.cap="Left pane of Figure 2.12"}     
fig2.12a = qplot(carat, data = diamonds, geom = "density", colour = color)    
fig2.12a
fig2.12a %>% VI()   
```


```{r fig2-12b, fig.cap="Right pane of Figure 2.12"}     
fig2.12b = qplot(carat, data = diamonds, geom = "histogram", fill = color)    
fig2.12b
fig2.12b %>% VI()   
```  

### bar charts for categorical variables   


```{r fig2-13a, fig.cap="Left pane of Figure 2.13"}     
fig2.13a  = qplot(color, data = diamonds, geom = "bar") #geom="bar" is the default    
fig2.13a
fig2.13a %>% VI()   
```


need to check...

```{r fig2-13b, fig.cap="Right pane of Figure 2.13"}     
fig2.13b = qplot(color, data = diamonds, geom = "bar", weight = carat)   
fig2.13b
fig2.13b = qplot(color, data = diamonds, geom = "bar", weight = carat) + scale_y_continuous("carat")    
fig2.13b
fig2.13b %>% VI()   
```  

## time series plots   

It looks like the data used in the next graph has been updated since the publication of @Wickham2009ggplot2 
```{r fig2-14a, fig.cap="Left pane of Figure 2.14"}     
fig2.14a = qplot(date, unemploy / pop, data = economics, geom = "line")    
fig2.14a
fig2.14a %>% VI()   
```  

## path plots   

```{r fig2-15b, fig.cap="Right pane of Figure 2.15"}     
year <- function(x) as.POSIXlt(x)$year + 1900    
fig2.15b = qplot(unemploy / pop, uempmed, data = economics, geom = "path", colour=year(date)) 
#+ scale_area() # no longer works
fig2.15b
fig2.15b %>% VI()   
```  


## facets is the ggplot term for trellis' panels   

The aspect ratio for the plot region is something that needs to be considered. I've manually adjusted the plotting window here so that he graph more closely matches that of @Wickham2009ggplot2 but it  is not an exact match.
```{r fig2-16a, fig.cap="Left side of Figure 2.16", fig.height=10, fig.width=4}     
fig2.16a = qplot(carat, data = diamonds, facets = color ~ ., geom = "histogram", binwidth = 0.1, 
xlim = c(0, 3))    
fig2.16a
fig2.16a %>% VI()   
```

```{r fig2-16b, fig.cap="Right side of Figure 2.16", fig.height=10, fig.width=4}
fig2.16b = qplot(carat, ..density.., data = diamonds, facets = color ~ ., geom = "histogram", binwidth = 0.1, xlim = c(0, 3))   
fig2.16b
fig2.16b %>% VI()   
```  

## rescaling of the axes   

```{r p26a, fig.cap="First graph on page 26 of "}
p26a = qplot(carat, price, data = dsmall, log = "xy")   
p26a
p26a %>% VI()   
```  

```{r fig3-6, fig.cap="Figure 3.6 of "}
fig3.6 = qplot(displ, hwy, data=mpg, facets =~ year) + geom_smooth()    
fig3.6
fig3.6 %>% VI()   
```  
