# Analysis of the Airquality2 data, using Ozone as the response variable and Month as the single grouping factor.
#### Prepared by Jonathan Godfrey   

```{r setup, purl=FALSE, include=FALSE}
library(BrailleR)
knitr::opts_chunk$set(dev=c("png", "pdf", "postscript", "svg"))
knitr::opts_chunk$set(echo=FALSE, comment="", fig.path="airquality2/Ozone.Month-", fig.width=7)
```

## Group summaries

```{r GroupSummary}
attach(airquality2)
Data.Mean <- tapply(Ozone, Month, mean, na.rm=TRUE)
Data.StDev <- tapply(Ozone, Month, sd, na.rm=TRUE)
nNonMissing <- function(x){
length(na.omit(x)) # length() includes NAs
}
Data.n <- tapply(Ozone, Month, nNonMissing)
Data.StdErr = Data.StDev/sqrt(Data.n)
detach(airquality2)
DataSummary = data.frame(names(Data.Mean), Data.Mean, Data.StDev, Data.n, Data.StdErr)
colnames(DataSummary) = c("Level", "Mean", "Standard deviation", "n", "Standard error")
```

```{r PrintSummary, results="asis", purl=FALSE}
kable(as.matrix(DataSummary), row.names=FALSE)
```  

```{r DataSummaryTex, purl=FALSE}
library(xtable)
ThisTexFile = "airquality2/Ozone.Month-GroupSummary.tex"
TabCapt = "Summary statistics for Ozone by level of Month"
print(xtable(DataSummary, caption=TabCapt, label="OzoneGroupSummary", digits=4, align="llrrrr"), include.rownames = FALSE, file = ThisTexFile)
```  

The ratio of the largest group standard deviation to the smallest is `r round(max(Data.StDev)/min(Data.StDev),2)`  

## Comparative boxplots  

 ```{r boxplots, fig.cap="Comparative boxplots"}  
 VI(boxplot(Ozone~Month, data=airquality2, ylab="Ozone", xlab="Month"))  
 ``` 
## Comparative dotplots

```{r dotplots, fig.cap="Comparative dotplots"}
with(airquality2,
VI(dotplot(Ozone~Month, xlab="Ozone", ylab="Month")))
``` 

## One-way Analysis of Variance

```{r OneWayANOVA1}
VI(
MyANOVA <- aov(Ozone~Month, data=airquality2)) 
```  

```{r ANOVA-TEX, purl=FALSE}
library(xtable)
ThisTexFile = "airquality2/Ozone-Month-ANOVA.tex"
TabCapt = "One-way ANOVA for Ozone with the group factor Month."
print(xtable(MyANOVA, caption=TabCapt, label="Ozone-Month-ANOVA", digits=4), file = ThisTexFile)
```  

```{r OneWayANOVA, fig.cap="Residual analysis"}
summary(MyANOVA)
par(mfrow=c(2,2))
plot(MyANOVA)
```

## Tests for homogeneity of Variance

```{r BartlettTest}
bartlett.test(Ozone~Month, data=airquality2)
```

```{r FlignerTest}
fligner.test(Ozone~Month, data=airquality2)
``` 

## Tukey Honestly Significant Difference test

```{r TukeyHSD, fig.cap="Plot of Tukey HSD"}
MyHSD <- TukeyHSD(MyANOVA, ordered=TRUE, conf.level=0.95)
VI(MyHSD)  

MyHSD
plot(MyHSD)
``` 

