# Analysis of the Airquality data, using Ozone as the response variable and Wind as the single predictor.
#### Prepared by  Jonathan Godfrey   

```{r setup2, purl=FALSE, include=FALSE}
library(BrailleR)
knitr::opts_chunk$set(dev=c("png", "pdf", "postscript", "svg"))
knitr::opts_chunk$set(echo=FALSE, comment="", fig.path="airquality/Ozone.Wind-", fig.width=7)
```

## Variable summaries

The response variable is Ozone and the predictor variable is Wind.

```{r VariableSummary}
attach(airquality)
SummaryResponse <- summary(na.omit(Ozone))
SummaryPredictor <- summary(na.omit(Wind))
SummaryTable <- cbind(Ozone=SummaryResponse,Wind=SummaryPredictor)
row.names(SummaryTable) <- c("Minimum","Lower Quartile", "Median","Mean","Upper Quartile","Maximum")
Sds <- c(sd(Ozone, na.rm=T),sd(Wind, na.rm=T))
Nas <- c(sum(is.na(Ozone)), sum(is.na(Wind)))
SummaryTable <- rbind(SummaryTable,"Standard Deviation"=Sds,"Missing Values"=Nas)
detach(airquality)
```

```{r PrintSummary, results="asis", purl=FALSE}
kable(t(SummaryTable), row.names=T, align=rep("c",8))
```  

```{r VariableSummaryTex, purl=FALSE}
library(xtable)
ThisTexFile = "airquality/Ozone.Wind-VariableSummary.tex"
TabCapt = "Summary statistics for variables Ozone and Wind."
print(xtable(t(SummaryTable), caption=TabCapt, label="Ozone-VariableSummary", digits=4, align="lrrrrrrrr"), include.rownames = FALSE, file = ThisTexFile)
```  

## Scatter Plot

```{r ScatterPlot, fig.cap="Scatter Plot"}
# Remove the missing values
completeCases <- complete.cases(Data[ResponseName])*complete.cases(Data[PredictorName])
assign(DataName, Data[completeCases==1,])

plot(Ozone~Wind, data=airquality, ylab=Ozone, xlab=Wind)
attach(airquality)
WhereXY(Ozone,Wind)
detach(airquality)
```  

## Linear regression

```{r SimpleLinMod}
Ozone.Wind.lm <- lm(Ozone~Wind, data=airquality)
VI( Ozone.Wind.lm )
VI(summary(Ozone.Wind.lm))
summary(Ozone.Wind.lm)
```

```{r FittedLinePlot}
plot(Ozone~Wind, data=airquality, ylab=Ozone, xlab=Wind)
abline(Ozone.Wind.lm)
```


```{r SimpleLinModResAnal, fig.cap="Residual analysis"}
par(mfrow=c(2,2))
plot(Ozone.Wind.lm)
```  

A separate html page showing the residual analysis and model validity checking for Ozone.Wind.lm is at [Ozone.Wind.lm.Validity.html](Ozone.Wind.lm.Validity.html)  

### One-way Analysis of Variance

```{r OneWayANOVA1}
anova(Ozone.Wind.lm)
```  

```{r SimpleLinMod-TEX, purl=FALSE}
ThisTexFile = "airquality/Ozone-Wind-lm.tex"
TabCapt = "Linear regression model for Ozone with the single Predictor Wind."
print(xtable(Ozone.Wind.lm, caption=TabCapt, label="Ozone-Wind-lm", digits=4), file = ThisTexFile)
```

```{r ANOVA-TEX, purl=FALSE}
ThisTexFile = "airquality/Ozone-Wind-anova.tex"
TabCapt = "Analysis of variance for the linear regression model having Ozone as the response and the Predictor Wind."
print(xtable(anova(Ozone.Wind.lm), caption=TabCapt, label="Ozone-Wind-lm", digits=4), file = ThisTexFile)
```  

