# Checking validity for the model "Ozone.Wind.lm" by way of standardised residuals, leverages, and Cook's distances
```{r GetVars, echo=FALSE}
Residuals=rstudent(Ozone.Wind.lm)
Fits= fitted(Ozone.Wind.lm)
Leverages= hatvalues(Ozone.Wind.lm)
Cooks= cooks.distance(Ozone.Wind.lm)
```    
#### Prepared by Jonathan Godfrey   

```{r setup, purl=FALSE, include=FALSE}
library(BrailleR)
opts_chunk$set(dev=c("png", "pdf", "postscript", "svg"))
opts_chunk$set(comment="", echo=FALSE, fig.path="Ozone.Wind.lm.Validity/Residuals-", fig.width=7)
```  

## Basic summary measures

```{r BasicSummaries}
Residuals.count = length(Residuals)
Residuals.unique = length(unique(Residuals))
Residuals.Nobs = sum(!is.na(Residuals))
Residuals.Nmiss = sum(is.na(Residuals))
Residuals.mean = mean(Residuals, na.rm = TRUE)
Residuals.tmean5 = mean(Residuals, trim =0.025, na.rm = TRUE)
Residuals.tmean10 = mean(Residuals, trim =0.05, na.rm = TRUE)
Residuals.IQR = IQR(Residuals, na.rm = TRUE)
Residuals.sd = sd(Residuals, na.rm = TRUE)
Residuals.var = var(Residuals, na.rm = TRUE)
Residuals.skew = moments::skewness(Residuals, na.rm = TRUE)
Residuals.kurt = moments::kurtosis(Residuals, na.rm = TRUE)
```

### Counts

`r Residuals.count` values in all, made up of
`r Residuals.unique` unique values,
`r Residuals.Nobs` observed, and
`r Residuals.Nmiss` missing values.   



### Measures of location

Data | all | 5% trimmed | 10% trimmed
----- | ------ | ----- | ------
Mean | `r Residuals.mean` | `r Residuals.tmean5` | `r Residuals.tmean10`

### Quantiles

```{r Quantiles1}
Quantiles=quantile(Residuals, na.rm=TRUE)
QList=c("Minimum", "Lower Quartile", "Median", "Upper Quartile", "Maximum")
Results=data.frame(Quantile=QList, Value=Quantiles[1:5])
```

```{r QuantilesPrint, eval=FALSE}
Results
```

```{r QuantilesKable, results="asis", purl=FALSE}
kable(Results, digits=4)
```

### Measures of spread

Measure | IQR | Standard deviation | Variance
-------- | ------ | -------- | ------
Value | `r Residuals.IQR` | `r Residuals.sd` | `r Residuals.var`   


## Basic univariate graphs    
### Histogram    

```{r Hist, fig.cap="The histogram", fig.height=5}    
VI(hist(Residuals, xlab="Residuals", main="Histogram of Residuals"))
```   


### Boxplot    

```{r Boxplot, fig.cap="The boxplot", fig.height=3.5}  
VI(boxplot(Residuals, horizontal=TRUE, main = "Boxplot of Residuals"))
```   


## Assessing normality

### Formal tests for normality

```{r NormalityTests}
library(nortest)
Results = matrix(0, nrow=6, ncol=2)
dimnames(Results) = list(c("Shapiro-Wilk", "Anderson-Darling", "Cramer-von Mises",
"Lilliefors (Kolmogorov-Smirnov)", "Pearson chi-square", "Shapiro-Francia"), c("Statistic", "P Value"))
 SW =shapiro.test(Residuals)
Results[1,] = c(SW$statistic, SW$p.value)
AD = ad.test(Residuals)
Results[2,] = c(AD$statistic, AD$p.value)
CV = cvm.test(Residuals)
Results[3,] = c(CV$statistic, CV$p.value)
LI = lillie.test(Residuals)
Results[4,] = c(LI$statistic, LI$p.value)
PE = pearson.test(Residuals)
Results[5,] = c(PE$statistic, PE$p.value)
SF = sf.test(Residuals)
Results[6,] = c(SF$statistic, SF$p.value)
```

```{r NormalityTestsPrint, eval=FALSE}
Results
```

```{r NormalityTestsKable, results="asis", purl=FALSE}
kable(Results, digits=c(4,4))
```   

### Normality plot    

```{r NormPlot, fig.cap="The normality plot", fig.height=7}  
qqnorm(Residuals, main = "Normality Plot for Residuals")
qqline(Residuals)
```   


## Formal tests of moments

```{r MomentsTests}
library(moments)
Results = matrix(0, nrow=2, ncol=3)
dimnames(Results)= list(c( "D'Agostino skewness", "Anscombe-Glynn kurtosis"),
 c("Statistic","Z",  "P Value"))
AG = moments::agostino.test(Residuals)
AN = moments::anscombe.test(Residuals)
Results[1,] = c(AG$statistic, AG$p.value)
Results[2,] = c(AN$statistic, AN$p.value)
```

```{r MomentsTestsPrint, eval=FALSE}
Results
```

```{r MomentsTests2, results="asis", purl=FALSE}
kable(Results, digits=c(4,3,4))
```   

## Regression diagnostic plots
### Standardised residuals

```{r Fits, fig.cap="Standardised residuals plotted against fitted values"}
plot(Fits, Residuals)
WhereXY(Fits, Residuals, yDist="normal")
```

```{r Order, fig.cap="Standardised residuals plotted against order"}
plot(Residuals)
WhereXY(1:length(Residuals), Residuals, yDist="normal")
```

```{r Lag1Resids, fig.cap="standardised residuals plotted against lagged residuals"}
n = length(Residuals)
plot(Residuals[-n], Residuals[-1], ylab= paste("Residuals 2 to", n), xlab=paste("Residuals 1 to",(n-1)))
WhereXY(Residuals[-n], Residuals[-1], xDist="normal")
```
The lag 1 autocorrelation of the standardised residuals is `r cor(Residuals[-n], Residuals[-1])`.

### Influence

```{r Leverages, fig.cap="Standardised residuals plotted against leverages"}
plot(Leverages, Residuals)
WhereXY(Leverages, Residuals, yDist="normal")
```

`r sum(Leverages>2*mean(Leverages))` points have excessive leverage.
`r sum(Cooks>1)` points have Cook's distances greater than one.

### Outliers and influential observations

```{r ListInfObs}
InflObs = data.frame(Ozone.Wind.lm$model, Fit=Fits, St.residual=Residuals, Leverage=Leverages, Cooks.distance=Cooks)[abs(Residuals)>2 | Cooks > 1 | Leverages > 2*mean(Leverages) , ]
```

```{r ListInfObsLatex, purl=FALSE}
library(xtable)
print(xtable(InflObs, caption="Listing of suspected outliers and influential observations.", label="InflObsOzone.Wind.lm", digits=4), file = "Ozone.Wind.lm.Validity/Influential.tex")
```

```{r ListInfObsKabled, results="asis", purl=FALSE}
kable(InflObs)
```   

